manuscript_figures

### 2024-08-07 scripts/manuscript_figures.Rmd
cleaned up manuscript script and output folders

nMDS or PCA on benthic community structure
```{r}
# Load in QUAD data and filter for invertebrates
invert.filt <- QUAD %>%
  filter(type %in% c("molluscs", "echinoderms", "crustacean"))

@@

# Calculate Bray-Curtis dissimilarity
m_com <- vegdist(numeric_data, method = "bray")

# Perform nMDS
nmds <- metaMDS(m_com, distance = "bray")

# Extract scores
scores <- as.data.frame(scores(nmds))

# Add row numbers as a unique identifier
scores$ID <- row.names(scores)

# Create a data frame with SITE and YEAR information
site_year_info <- wide_data[rowSums(wide_data[, c("crustacean", "echinoderms", "molluscs")]) > 0, c("SITE", "YEAR")]
site_year_info$ID <- row.names(site_year_info)

# Merge SITE and YEAR information into scores
scores <- merge(scores, site_year_info, by = "ID")

# Define custom shape values for more than 8 categories
custom_shape_values <- c(0, 1, 2, 3, 4, 5, 6, 23)

# Use these custom shapes in your plot
nmds_plot <- ggplot(scores, aes(x = NMDS1, y = NMDS2, color = SITE, shape = as.factor(YEAR))) +
  geom_point(size = 8) +
  #stat_ellipse(aes(group = YEAR), level = 0.95, linetype = 2, size = 1, alpha = 0.5) + # Add ellipses
  labs(x = "NMDS1", y = "NMDS2", color = "Site", shape = "Year") +
  scale_color_manual(values = c("Fort Wetherill" = "grey", "King's Beach" = "black")) +
  scale_shape_manual(values = custom_shape_values) +  # Specify the custom shapes
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18)
  );nmds_plot

# Save the plot to a file
ggsave("output/figures/nmds.plot.png", plot = nmds_plot, width = 15, height = 10)

@@

PERMANOVA and SIMPER analysis
```{r}
# Ensure only numeric columns are selected for vegdist
community_data <- wide_data[, c("crustacean", "echinoderms", "molluscs")]

@@

# Create environmental data frame
environmental_data <- wide_data[non_zero_rows, c("TRANSECT", "YEAR", "SITE")]

# Convert YEAR to factor
environmental_data$YEAR <- as.factor(environmental_data$YEAR)

# Perform PERMANOVA
permanova_result <- adonis2(community_data ~ SITE + YEAR, data = environmental_data, permutations = 999, method = "bray")

# View the results
print(permanova_result)

# Create dataframe from PERMANOVA results
permanova_df <- data.frame(
  Factor = c("SITE", "YEAR", "Residual", "Total"),
  Df = c(1, 7, 44, 52),
  SumOfSqs = c(0.5649, 5.0492, 8.7778, 14.3919),
  R2 = c(0.03925, 0.35084, 0.60991, 1.00000),
  F = c(2.8317, 3.6157, NA, NA),
  Pr_F = c(0.050, 0.001, NA, NA)  # Changed column name to Pr_F
)

# Create flextable object
flex_perm <- flextable(permanova_df)

# Customize the flextable
flex_perm <- flex_perm %>%
  set_header_labels(
    Factor = "Factor",
    Df = "Df",
    SumOfSqs = "Sum of Squares",
    R2 = "R2",
    F = "F",
    Pr_F = "Pr(>F)"  # Changed to match the new column name
  ) %>%
  colformat_double(digits = 4) %>%
  bold(i = ~ Pr_F <= 0.05, j = "Pr_F") %>%  # Bold values if Pr_F is 0.05 or lower
  add_header_lines(values = c(
    "Permutation test for adonis under reduced model",
    "Number of permutations: 999"
  )) %>%
  autofit() %>%
  theme_box()

# Save flextable as PNG
png_file <- "output/model.result.tables/permanova.png"
save_as_image(flex_perm, path = png_file)

# Perform SIMPER analysis
simper_result <- simper(community_data, group = environmental_data$SITE, permutations = 999)

# View summary of results
summary(simper_result)

# Create dataframe from SIMPER results
simp <- data.frame(
  Species = rownames(summary(simper_result)$`Fort Wetherill_King's Beach`),
  Average = summary(simper_result)$`Fort Wetherill_King's Beach`$average,
  Ratio = summary(simper_result)$`Fort Wetherill_King's Beach`$ratio,
  Fort_Wetherill = summary(simper_result)$`Fort Wetherill_King's Beach`$ava,
  Kings_Beach = summary(simper_result)$`Fort Wetherill_King's Beach`$avb,
  Cumulative_Sum = summary(simper_result)$`Fort Wetherill_King's Beach`$cumsum,
  p_value = summary(simper_result)$`Fort Wetherill_King's Beach`$p
)

# Create flextable object
flex_simp <- flextable(simp)

# Customize the flextable
flex_simp <- flex_simp %>%
  set_header_labels(
    Species = "Species",
    Average = "Avg. Contribution",
    Ratio = "Ratio",
    Fort_Wetherill = "Fort Wetherill Abundance",
    Kings_Beach = "King's Beach Abundance",
    Cumulative_Sum = "Cumulative Sum",
    p_value = "p-value"
  ) %>%
  autofit() %>%
  theme_box()

# Save flextable as PNG
png_file <- "output/model.result.tables/simper_results.png"
save_as_image(flex_simp, path = png_file)

# Extract the relevant information from the SIMPER results
simper_data <- data.frame(
  Species = rownames(summary(simper_result)$`Fort Wetherill_King's Beach`),
  Contribution = summary(simper_result)$`Fort Wetherill_King's Beach`$average,
  Fort_Wetherill = summary(simper_result)$`Fort Wetherill_King's Beach`$ava,
  Kings_Beach = summary(simper_result)$`Fort Wetherill_King's Beach`$avb
)

# Calculate the percentage contribution
simper_data$Percentage <- simper_data$Contribution / sum(simper_data$Contribution) * 100

# Reshape the data for plotting
simper_long <- tidyr::pivot_longer(simper_data, 
                                   cols = c(Fort_Wetherill, Kings_Beach),
                                   names_to = "Site", 
                                   values_to = "Abundance")

# Create the plot
simper_plot <- ggplot(simper_long, aes(x = Species, y = Abundance, fill = Site)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.1f%%", Percentage[match(Species, simper_data$Species)])),
            position = position_dodge(width = 0.9),
            vjust = -0.5, 
            size = 3) +
  labs(title = "Species Contributions to Differences Between Sites",
       subtitle = "Based on SIMPER analysis",
       y = "Average Abundance",
       x = "Species") +
  scale_fill_manual(values = c("Fort_Wetherill" = "grey", "Kings_Beach" = "black")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.title = element_blank());simper_plot

# Save the plot to a file
ggsave("output/figures/simper.plot.png", plot = simper_plot, width = 15, height = 10)

PERMANOVA results:
adonis2(formula = community_data ~ SITE + YEAR, data = environmental_data, permutations = 999, method = "bray")
         Df SumOfSqs      R2      F Pr(>F)    
SITE      1   0.5649 0.03925 2.8317  0.043 *  
YEAR      7   5.0492 0.35084 3.6157  0.001 ***
Residual 44   8.7778 0.60991                  
Total    52  14.3919 1.00000

SIMPER results:
ontrast: Fort Wetherill_King's Beach

            average      sd   ratio     ava     avb cumsum     p
molluscs    0.58750 0.29784 1.97250 5.57600 5.84300  0.863 0.576
echinoderms 0.07140 0.12811 0.55760 0.21400 0.34300  0.968 0.547
crustacean  0.02160 0.05018 0.43030 0.21900 0.09300  1.000 0.526
Permutation: free
Number of permutations: 999

# Figure for functional group change in abundance over time using invert count data
```{r}
# Check for normality of residuals
# Log transform the abundance data
invert.count$log_abundance <- log(invert.count$abundance + 1.1)

# Check for normality of transformed data
anova_model <- aov(log_abundance ~ type * YEAR, data = invert.count)
residuals <- residuals(anova_model)

@@

# Perform Kruskal-Wallis test for interaction type and YEAR
interaction_data <- invert.count %>%
  mutate(interaction = interaction(type, YEAR))

kruskal_test_interaction <- kruskal.test(abundance ~ interaction, data = interaction_data)
print(kruskal_test_interaction)

@@

# Assuming invert.count is your dataframe and abundance_g_per_m2 is already calculated
total_abundance <- invert.count %>%
  group_by(YEAR) %>%
  summarise(total_abundance_g_per_m2 = sum(abundance))

@@

data:  abundance by interaction
Kruskal-Wallis chi-squared = 104.02, df = 23, p-value = 2.83e-12

>>>>

Invertebrate community structure calculations
```{r}
# Load in QUAD data and filter for invertebrates
invert.filt <- QUAD %>%
  filter(type %in% c("molluscs", "echinoderms", "crustacean"))

@@

@@

Figure and calculations for functional group change in relative abundance over time using invert count data
```{r}
# Ensure only numeric columns are selected for vegdist
community_data <- wide_data[, c("crustacean", "echinoderms", "molluscs")]

@@

# Calculate total abundance 
total_abundance <- invert.count %>%
  group_by(YEAR) %>%
  summarise(total_abundance_g_per_m2 = sum(abundance))

# Check for normality of residuals
# Log transform the abundance data
invert.abun$log_rel_abundance <- log(invert.abun$abundance + 1.1)

# Check for normality of transformed data
anova_model <- aov(log_rel_abundance ~ type * YEAR, data = invert.abun)
residuals <- residuals(anova_model)

@@

# Perform Kruskal-Wallis test for interaction type and YEAR
interaction_data <- invert.abun %>%
  mutate(interaction = interaction(type, YEAR))

kruskal_test_interaction <- kruskal.test(log_rel_abundance ~ interaction, data = interaction_data)
print(kruskal_test_interaction)

@@

@@

data:  log_rel_abundance by interaction
Kruskal-Wallis chi-squared = 104.02, df = 23, p-value = 2.83e-12
